---
title: "Analyzing the Effects of NAFTA on US Trade with Canada and Mexico"
author: "C.J. Laws, Zack Roder, Graham Shank"
date: "4/22/2021"
output: ioslides_presentation
---

```{r setup, include=FALSE}
library(astsa)
library(readxl)
library(priceR)

data <- read_excel("C:\\Users\\Zack\\Documents\\Spring 2021\\MATH 333\\project\\EXPORTS_IMPORTS_MEX_CAN.xls")

data$CAN_EXPORTS <- adjust_for_inflation(data$CAN_EXPORTS, data$observation_date, "US", to_date=2019)
data$CAN_IMPORTS <- adjust_for_inflation(data$CAN_IMPORTS, data$observation_date, "US", to_date=2019)

data$MEX_EXPORTS <- adjust_for_inflation(data$MEX_EXPORTS, data$observation_date, "US", to_date=2019)
data$MEX_IMPORTS <- adjust_for_inflation(data$MEX_IMPORTS, data$observation_date, "US", to_date=2019)

#use millions of dollars as unit
data$CAN_EXPORTS <- data$CAN_EXPORTS / 1000000 
data$CAN_IMPORTS <- data$CAN_IMPORTS / 1000000
data$MEX_EXPORTS <- data$MEX_EXPORTS / 1000000
data$MEX_IMPORTS <- data$MEX_IMPORTS / 1000000

#convert to ts object
CAN_EXPORTS <- ts(data$CAN_EXPORTS, start=1980, freq=12)
CAN_IMPORTS <- ts(data$CAN_IMPORTS, start=1980, freq=12)
MEX_EXPORTS <- ts(data$MEX_EXPORTS, start=1980, freq=12)
MEX_IMPORTS <- ts(data$MEX_IMPORTS, start=1980, freq=12)
```

## Introduction and Project Goals
- NAFTA, which established a free-trade zone on the North American Continent, went into effect in January 1994.
- To understand the effects of NAFTA on US trade with Canada and Mexico, we hope to develop a model to create a forecast with pre-NAFTA data.
- We could compare the forecasted trade numbers to the actual, post-NAFTA data to see if NAFTA significantly increased US trade with Mexico and Canada.
![](NAFTA_logo.svg){width=30%}

## Our Data
- To explore this question, we located data from FRED. We found 4 series that are useful in quantifying trade activity.
  + Exports: Value Goods for Canada (TEXVA01CAM667S)
  + Imports: Value Goods for Canada (TIMVA01CAM667S)
  + Exports: Value Goods for Mexico (TEXVA01MXM667S)
  + Imports: Value Goods for Mexico (TIMVA01MXM667S)
- The units for each series are US Dollars, Seasonally Adjusted. We retrieved data for 1980 through 2019.
- We first adjusted the series for inflation, such that the units are now 2019 US Dollars. This may, perhaps, help with some irregularities in the series. Even if it does not, it is helpful in interpreting the series over time.

## Our Data
Here is a preliminary plot of each of the four series:
```{r 1, echo = FALSE}
par(mfrow=c(2,2))
tsplot(CAN_EXPORTS, main="Monthly US Exports Canada, 1980-2019", ylab= "2019 US Dollars, Millions", xlab="Year")
tsplot(CAN_IMPORTS, main="Monthly US Imports from Canada, 1980-2019", ylab= "2019 US Dollars, Millions", xlab="Year")
tsplot(MEX_EXPORTS, main="Monthly US Exports to Mexico, 1980-2019", ylab= "2019 US Dollars, Millions", xlab="Year")
tsplot(MEX_IMPORTS, main="Monthly US Imports from Mexico, 1980-2019", ylab= "2019 US Dollars, Millions", xlab="Year")
```

## Our Data
To make one series, we sum together the four series. This combined series can now be interpreted as the monthly total trade with both Mexico and Canada. Also, since we are particularly interested in modeling the data between 1980 and 1994, we will splice it.
```{r 2, echo=FALSE, fig.width=6, fig.height=4, fig.align="center"}
TOTAL_TRADE = CAN_EXPORTS + CAN_IMPORTS + MEX_EXPORTS + MEX_EXPORTS
TOTAL_TRADE = ts(TOTAL_TRADE[1:168], start=1980, freq=12)
tsplot(TOTAL_TRADE, main="Monthly US Total Trade with Canada and Mexico, 1980-1994", ylab="2019 US Dollars, Millions", xlab="Year")
```

##Preliminary Analysis
We see an obvious upward trend, there is no immediately apparent seasonality. As the ACF confirms, we will first need to coerce this data into being stationary before we can build our model and forecast.
