---
title: "MATH 333 Project"
author: "C.J. Laws, Zack Roder, Graham Shank"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
library(priceR)
```

## Introduction and Motivation
In our project, we are interested in exploring the effects of NAFTA on US imports & exports to both Canada and Mexico. NAFTA, which went into effect in January 1994, established a free-trade zone on the North American continent.

To understand the effects of NAFTA on US trade with Canada and Mexico, we hope to develop a model of US exports/imports with the two nations using data prior to 1994. With a model, we could compare a forecast of imports/exports with the real post-1994 data to see if how significant of an effect NAFTA has had on North American trade.

We located our data on https://fred.stlouisfed.org. Units are US Dollars, Seasonally Adjusted.


## Data Cleaning and Initial Analysis
First, we clean the data so that we can load it into a ts object.

```{r}
library(readxl)
library(astsa)

#read data from excel
data <- read_excel("C:\\Users\\Zack\\Documents\\Spring 2021\\MATH 333\\project\\EXPORTS_IMPORTS_MEX_CAN.xls")
data <- data.frame(data$CAN_EXPORTS, data$CAN_IMPORTS, data$MEX_EXPORTS, data$MEX_IMPORTS)

#convert to ts object
data_ts <- ts(data, start=1980, freq=12)

tsplot(data_ts)
```

After plotting the data, we can see that all series have a strong, upward trends. Somehow, we need to make this data stationary. First, it may be helpful to adjust the data for inflation (since it's currently in nominal dollars) to remove at least some of the irregularity and better understand the dollar values. Using the priceR package, we can do this easily. We'll convert each to 2019 dollars (last year of series).

```{r}
data <- read_excel("C:\\Users\\Zack\\Documents\\Spring 2021\\MATH 333\\project\\EXPORTS_IMPORTS_MEX_CAN.xls")

data$CAN_EXPORTS <- adjust_for_inflation(data$CAN_EXPORTS, data$observation_date, "US", to_date=2019)
data$CAN_IMPORTS <- adjust_for_inflation(data$CAN_IMPORTS, data$observation_date, "US", to_date=2019)

data$MEX_EXPORTS <- adjust_for_inflation(data$MEX_EXPORTS, data$observation_date, "US", to_date=2019)
data$MEX_IMPORTS <- adjust_for_inflation(data$MEX_IMPORTS, data$observation_date, "US", to_date=2019)



#convert to ts object
CAN_EXPORTS <- ts(data$CAN_EXPORTS, start=1980, freq=12)
CAN_IMPORTS <- ts(data$CAN_IMPORTS, start=1980, freq=12)
MEX_EXPORTS <- ts(data$MEX_EXPORTS, start=1980, freq=12)
MEX_IMPORTS <- ts(data$MEX_IMPORTS, start=1980, freq=12)

data <- data.frame(data$CAN_EXPORTS, data$CAN_IMPORTS, data$MEX_EXPORTS, data$MEX_IMPORTS)

data <- ts(data, start=1980, freq=12)

tsplot(data_ts)
```

Let's see what the ACF currently looks like.
```{r}
acf1(CAN_EXPORTS)
acf1(CAN_IMPORTS)
acf1(MEX_EXPORTS)
acf1(MEX_IMPORTS)
```
As we can see, this is not even close to being stationary. First, we'll try differencing.

```{r}
CAN_EXPORTS_diff <- diff(CAN_EXPORTS)
CAN_IMPORTS_diff <- diff(CAN_IMPORTS)
MEX_EXPORTS_diff <- diff(MEX_EXPORTS)
MEX_IMPORTS_diff <- diff(MEX_IMPORTS)

acf1(CAN_EXPORTS_diff)
acf1(CAN_IMPORTS_diff)
acf1(MEX_EXPORTS_diff)
acf1(MEX_IMPORTS_diff)

tsplot(CAN_EXPORTS_diff)
tsplot(CAN_IMPORTS_diff)
tsplot(MEX_EXPORTS_diff)
tsplot(MEX_IMPORTS_diff)

```

Detrending:

```{r}
fit <- lm(CAN_EXPORTS ~ time(CAN_EXPORTS), na.action=NULL)
tsplot(resid(fit))

```