---
title: "MATH 333 Project"
author: "C.J. Laws, Zack Roder, Graham Shank"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
library(readxl)
library(priceR)
library(forecast)
```

## Introduction and Motivation
In our project, we are interested in exploring the effects of NAFTA on US imports & exports to both Canada and Mexico. NAFTA, which went into effect in January 1994, established a free-trade zone on the North American continent.

To understand the effects of NAFTA on US trade with Canada and Mexico, we hope to develop a model of US exports/imports with the two nations using data prior to 1994. With a model, we could compare a forecast of imports/exports with the real post-1994 data to see if how significant of an effect NAFTA has had on North American trade.

We located our data on https://fred.stlouisfed.org. Units are US Dollars, Seasonally Adjusted.


## Data Cleaning and Initial Analysis
First, we clean the data so that we can load it into a ts object. First, it may be helpful to adjust the data for inflation (since it's currently in nominal dollars) to remove at least some of the irregularity and better understand the dollar values. Using the priceR package, we can do this easily. We'll convert each to 2019 dollars (last year of series). We then plot the data. After plotting the data, we can see that all series have a strong, upward trends. Somehow, we need to make these data stationary. 

```{r, echo=FALSE, results='hide'}
data <- read_excel("C:\\Users\\Zack\\Documents\\Spring 2021\\MATH 333\\project\\EXPORTS_IMPORTS_MEX_CAN.xls")

data$CAN_EXPORTS <- adjust_for_inflation(data$CAN_EXPORTS, data$observation_date, "US", to_date=2019)
data$CAN_IMPORTS <- adjust_for_inflation(data$CAN_IMPORTS, data$observation_date, "US", to_date=2019)

data$MEX_EXPORTS <- adjust_for_inflation(data$MEX_EXPORTS, data$observation_date, "US", to_date=2019)
data$MEX_IMPORTS <- adjust_for_inflation(data$MEX_IMPORTS, data$observation_date, "US", to_date=2019)

#use millions of dollars as unit
data$CAN_EXPORTS <- data$CAN_EXPORTS / 1000000 
data$CAN_IMPORTS <- data$CAN_IMPORTS / 1000000
data$MEX_EXPORTS <- data$MEX_EXPORTS / 1000000
data$MEX_IMPORTS <- data$MEX_IMPORTS / 1000000

#convert to ts object
CAN_EXPORTS <- ts(data$CAN_EXPORTS, start=1980, freq=12)
CAN_IMPORTS <- ts(data$CAN_IMPORTS, start=1980, freq=12)
MEX_EXPORTS <- ts(data$MEX_EXPORTS, start=1980, freq=12)
MEX_IMPORTS <- ts(data$MEX_IMPORTS, start=1980, freq=12)

par(mfrow=c(2,2))

tsplot(CAN_EXPORTS, main="Monthly US Exports to Canada, 1980-2019", ylab= "2019 US Dollars, Millions (inflation adjusted)", xlab="Year" )
tsplot(CAN_IMPORTS, main="Monthly US Imports from Canada, 1980-2019", ylab= "2019 US Dollars, Millions (inflation adjusted)", xlab="Year")
tsplot(MEX_EXPORTS, main="Monthly US Exports to Mexico, 1980-2019", ylab= "2019 US Dollars, Millions (inflation adjusted)", xlab="Year" )
tsplot(MEX_IMPORTS, main="Monthly US Imports from Mexico, 1980-2019", ylab= "2019 US Dollars, Millions (inflation adjusted)", xlab="Year")
```

So that we have just one series to analyze, we can sum these four series to get a new series: US Total Trade with Mexico and Canada, Monthly. Furthermore, since we are particularly interested in pre-NAFTA data, we will splice the series.
```{r, echo=FALSE, results='hide'}
TOTAL_TRADE_EXT = CAN_EXPORTS + CAN_IMPORTS + MEX_EXPORTS + MEX_IMPORTS
TOTAL_TRADE = ts(TOTAL_TRADE_EXT[1:168], start=1980, freq=12)
tsplot(TOTAL_TRADE, main="Monthly US Total Trade with Canada and Mexico, 1980-1994", ylab="2019 US Dollars, Millions", xlab="Year")
```

As we can see, this is not even close to being stationary. Hopefully, the difference and log transformation will coerce the data into being approximately stationary.
```{r, results='hide'}
tsplot(diff(log(TOTAL_TRADE)))
acf2(diff(log(TOTAL_TRADE)))
```

The logged series with first difference applied appears approximately stationary, so we can proceed with building our model.

## Building a SARIMA model
To look for seasonal components for our model, we need to also account for seasonal differences in the series. Since this is monthly, we will take the lag 12 difference as well.

```{r, echo=FALSE, results='hide'}
tsplot(diff(diff(log(TOTAL_TRADE), lag=12)), col=4, ylab=expression(nabla~nabla[12]~log(TOTAL_TRADE)))
```

By evaluating the ACF/PACF, we should be able to determine the SARIMA components.

```{r, results='hide'}
acf2(diff(diff(log(TOTAL_TRADE), lag=12)))
```

To estimate the non-seasonal order of the SARIMA model, we inspect the first few lags. It appears that the PACF cuts off after lag 2. This suggests AR(2) within the season. Including a difference order, this gives us an estimate of ARIMA(2,1,0).

To estimate the seasonal order of the SARIMA model, we inspect the lags at 12s (s=1,2,...). At lag 1 (12 months), it appears that they both tail off, indicating ARMA. We will need to find the SAR and SMA orders through trial and error.

We will first try an $ARIMA(2,1,0) \times (1,1,1)_{12}$ model.

```{r}
sarima(log(TOTAL_TRADE), 2, 1, 0, 1, 1, 1, 12)
```

The residuals for this model look okay, but we will see if adding another AR component within the season helps.

```{r}
sarima(log(TOTAL_TRADE), 2, 1, 0, 2, 1, 1, 12)
```

Overall, this $ARIMA(2,1,0) \times (2,1,1)_{12}$ model appears pretty good. The ACF values for the residuals are all insignificant, and all but one of the Ljung-Box statistics are insignificant. The Q-Q plot appears reasonably normal aswell. Hence, we will proceed with this model for our forecast.

## Forecast
Using the model we developed, we can forecast values from 1994-2000 and compare those projected values to the actual values. This should be able to reveal noticeable increases in trade activity that occurred immediately following the implementation of NAFTA. Using the 1980-1994 data, we project out 60 months and compare that to the actual data points.

```{r}
mod <- sarima.for(log(TOTAL_TRADE), 60, 2, 1, 0, 2, 1, 1, 12)
lines(log(ts(TOTAL_TRADE_EXT[168:length(TOTAL_TRADE_EXT)], start=1994, freq=12)), type="o", col="blue")
```

As we can see, nearly all of the actual data points are outside of the 80% confidence interval of the forecast (darker shaded region). This seems to reveal that NAFTA likely did have a somewhat noticeable effect on North American trade during the 5 years immediately following its implementation.


## Other Methods We Tried

### Detrending
Because of the clear upward trend present in the series, we though that detrending (then developing an ARIMA model on the residuals) might be appropriate. However, we had mixed results.

First, we detrended the data:

```{r}
trend <- lm(TOTAL_TRADE ~  time(TOTAL_TRADE), na.action = NULL)
summary(trend)
tsplot(TOTAL_TRADE, main="Monthly US Total Trade with Canada and Mexico, 1980-1994", ylab="2019 US Dollars, Millions", xlab="Year")
lines(fitted(trend), col="red")
```

The linear model for the trend looks pretty solid, with significant coefficients. Next, we look at the plot of the residuals:

```{r}
tsplot(resid(trend))
```

This is obviously non-stationary, so we need to look at the difference.

```{r}
tsplot(diff(resid(trend)))
acf2(diff(resid(trend)))

decompose(resid(trend))
```