---
title: "To do the simulation portion"
author: "Please answer here the simulation questions"
output: html_document
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In each of the followign questions, I want you to show the **R codes** and the **graphes** when required. Please Submit the **HTML** file, no rmd file is needed.  

Before anything, we may need to load the used packages, ex: the astsa package.

```{r}
library(astsa)
```

##### Problem 1- b: Simulate and plot a white noise with mean zero and standard deviation equal to 5.  

```{r 1b}
w = rnorm(n=500, mean=0, sd=5)
tsplot(w)
acf1(w)
```

##### Problem 2- a: Simulate and plot the process Vt
```{r 2a}
w = rnorm(502) #extra samples
v = filter(w, filter=rep(1/3,3), method="convolution", sides=2)
par(mfrow=c(2,1))
tsplot(w, main = "White noise")
tsplot(v, main = "Moving average")

```

##### Problem 3- c: Simulate and plot **any path** of the process Xt. 
Plotting $x_t$ with drift $\delta = 0$:
```{r 3c}
w = rnorm(500)
x = cumsum(w)
tsplot(x, main="Random walk w/o drift")
acf1(x)
```

##### Problem 6- a: Simulate and plot any path of an AR(1) process and plot the autocorrelation function. 
```{r 6a}
w = rnorm(502) #extra samples
x = filter(w, filter=c(0.5), method="recursive")[-(1:2)] #choose phi = 0.5
tsplot(x, main="AR(1)")
```

